Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems.

### 2023-11-27 article.qmd
Add imports as part of poincare code figure
SimBio figure

```{python}
# | echo: false
from poincare import Derivative, System, Variable, initial

@@

:::: {#fig-code layout-nrow=3}

@@

For reaction networks,
we are usually interested in first-order differential equations
relating the rate of change of the network's species.

To facilitate defining these models,
SimBio defines `Species`,
which comprises of a `poincare.Variable` and an stoichiometry,
and `Reaction`s,
which transform reactants into products taking into account the stoichiometry.

```{python}
from simbio import Compartment, Species, Reaction, initial

class Model(Compartment):
    """2A -> B"""

    A: Species = initial(default=1)
    B: Species = initial(default=0)
    r = Reaction(reactants=[2 * A], products=[B], rate_law=1)

@@

It also includes a `MassAction` variant,
which automatically adds the reactants as part of the rate.
`MassAction(reactants=[2 * A], product=[B], rate=1)`
is equivalent to
`Reaction(reactants=[2 * A], product=[B], rate=1 * A**2)`,
and many predefined commonly used reactions reactions.

SimBio also supports importing from and exporting to SBML,
or directly from BioModels:

```python
from simbio import Simulator
from simbio.io import biomodels, sbml

model = sbml.load("repressilator.sbml")  # from existing SBML file
model = biomodels.load_model("BIOMD12")  # from BioModels
Simulator(model).solve(...)
```

>>>>

:::: {#fig-poincare layout-nrow=4}

```{python}
from poincare import Derivative, System, Variable, initial

@@

@@

:::: {#fig-simbio layout-nrow=3}

```{python}
from simbio import Compartment, MassAction, Species, Reaction, initial

@@

::: {#fig-reaction layout-ncol=2}
```{python}
class Model(Compartment):
    """2A -> B"""

    A: Species = initial(default=1)
    B: Species = initial(default=0)
    r = Reaction(
        reactants=[2 * A],
        products=[B],
        rate_law=1,
    )

$$
\begin{align}
&\begin{cases}
    \dot{A} = -2 \\
    \dot{B} = +1
\end{cases}
\\\\
&\begin{cases}
    A(0) &= 1 \\
    B(0) &= 0
\end{cases}
\end{align}
$$

Reaction system.
:::

::: {#fig-mass-action layout-ncol=2}
```{python}
class Model(Compartment):
    """2A -> B"""

    A: Species = initial(default=1)
    B: Species = initial(default=0)
    r = MassAction(
        reactants=[2 * A],
        products=[B],
        rate=1,
    )

$$
\begin{align}
&\begin{cases}
    \dot{A} = -2 A^2 \\
    \dot{B} = +1 A^2
\end{cases}
\\\\
&\begin{cases}
    A(0) &= 1 \\
    B(0) &= 0
\end{cases}
\end{align}
$$

Mass-Action system.
:::

::::

For reaction networks,
we are interested in first-order differential equations
relating the rate of change of the species.
To facilitate defining these models,
SimBio defines `Species`,
which comprises of a `poincare.Variable` and an stoichiometry,
and `Reaction`,
which transform reactants into products taking into account the stoichiometry (@fig-reaction).
It also includes `MassAction`, a `Reaction` subclass,
which automatically adds the reactants as part of the rate law (@fig-mass-action).
Several commonly used reactions are predefined as `MassAction` subclasses.
Additionally, SimBio supports importing from and exporting to SBML,
or directly from BioModels (@fig-simbio-io).

::: {#fig-simbio-io layout-ncol=1}
```{python}
from simbio.io import biomodels, sbml

model = sbml.load("repressilator.sbml")  # from existing SBML file
model = biomodels.load_model("BIOMD12")  # from BioModels
```

Creation of a model from a local SBML file or one uploaded to BioModels.
:::
